---
title: "Example R and Stata code for a Mendelian randomization analysis"
author: "Sanderson et al."
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
  word_document:
    reference_docx: "custom-reference.docx"
bibliography: references.bib 
---

```{r include=FALSE, message=FALSE}
options(width = 100)

suppressPackageStartupMessages({
  
  if (!requireNamespace("remotes", quietly = TRUE))
    install.packages("remotes")
  
  if (!requireNamespace("tidyverse", quietly = TRUE))
    install.packages("tidyverse")
  
  if (!requireNamespace("MendelianRandomization", quietly = TRUE))
    install.packages("MendelianRandomization")

  if (!requireNamespace("sessioninfo", quietly = TRUE)) 
    install.packages("sessioninfo")

})
```

<!-- Move TOC to RHS -->
```{css toc-content, echo = FALSE}
#TOC {
  right: 270px;
  margin: 20px 0px 25px 0px;
}

.main-container {
    margin-left: 200px;
}
```

# Example R and Stata code to perform a multivariable Mendelian randomization (MVMR) analysis

## R code

* Read in the data

```{r}
suppressPackageStartupMessages({
  library(tidyverse)
  library(haven)
  library(remotes)
})
dat <- read_dta("https://raw.github.com/remlapmot/mrrobust/master/dodata.dta")
dat <- dat %>% filter(ldlcp2 < 1e-8)
```

### Example code using the MendelianRandomization package

* Install the package and load it into memory

```{r}
library(MendelianRandomization)
```

* Convert our data frame to the required class

```{r}
datfmt <- mr_mvinput(
  bx = as.matrix(cbind(dat$ldlcbeta, dat$hdlcbeta, dat$tgbeta)),
  bxse = as.matrix(cbind(dat$ldlcse, dat$hdlcse, dat$tgse)),
  by = dat$chdbeta,
  byse = dat$chdse,
  exposure = "exposure",
  outcome = "outcome",
  snps = "snp",
  effect_allele = dat$a1,
  other_allele = dat$a2,
  eaf = NA
)
```

* Fit an MVMR/MVIVW model

```{r}
mvivwfit <- mr_mvivw(datfmt)
mvivwfit
```

* Fit an MVMR-Egger model

```{r}
mvmreggerfit <- mr_mvegger(datfmt)
mvmreggerfit
```

### Example code using the MVMR and RMVMR packages

* Install the package and load it into memory

```{r}
if (!requireNamespace("MVMR", quietly = TRUE)) {
  remotes::install_github("WSpiller/MVMR")
}
library(MVMR)
```

* Create a data object of the required structure

```{r}
r_input <- format_mvmr(
  BXGs = dat[,c("ldlcbeta","hdlcbeta","tgbeta")],
  BYG = dat$chdbeta,
  seBXGs = dat[,c("ldlcse","hdlcse","tgse")],
  seBYG = dat$chdse,
  RSID = dat$rsid
)
```

* Fit an MVMR model

```{r cache=TRUE}
mvmrfit <- ivw_mvmr(r_input)
```

* Conditional F-statistics for instrument strength [@sanderson-statsmed-2021]

```{r cache=TRUE}
strength_mvmr(r_input)
```

* Fit a radial MVMR model

```{r cache=TRUE}
if (!requireNamespace("RMVMR", quietly = TRUE)) {
  remotes::install_github("WSpiller/RMVMR")
}
library(RMVMR)

rmvmr_input <- mrmvinput_to_rmvmr_format(datfmt)

rmvmr_fit <- ivw_rmvmr(rmvmr_input, summary = TRUE)
rmvmr_fit$coef
```

* Plot the radial MVMR models

```{r cache=TRUE}
plt_rmvmr <- plot_rmvmr(rmvmr_input, rmvmr = rmvmr_fit)
plt_rmvmr$p1
plt_rmvmr$p2
```

* Heterogeneity statistics

```{r cache=TRUE}
pleio_rmvmr <- pleiotropy_rmvmr(rmvmr_input, rmvmr = rmvmr_fit)
pleio_rmvmr$gq
head(pleio_rmvmr$qdat)
```

* Conditional F-statistics for instrument strength [@sanderson-statsmed-2021]

```{r cache=TRUE}
str_rmvmr <- strength_rmvmr(rmvmr_input)
str_rmvmr$plot[[2]]
str_rmvmr$qstat[[2]]
```

## Stata code

* Load the Statamarkdown package to enable Stata code chunks in an R Markdown file

```{r, message=FALSE}
if (!requireNamespace("Statamarkdown", quietly = TRUE)) {
  remotes::install_github("Hemken/Statamarkdown")
}
library(Statamarkdown)
```

* Read in the data and create an indicator variable to select observations with *p*-value between the genotype and LDL-C < 10<sup>-8</sup>

```{stata, collectcode=TRUE}
use https://raw.github.com/remlapmot/mrrobust/master/dodata, clear
gen byte sel1 = (ldlcp2 < 1e-8)
```

### Example code using the mrrobust package

* Install the mrrobust package using the github package

```{stata results=FALSE, eval=FALSE}
// Note: output suppressed
net install mrrobust, from("https://raw.github.com/remlapmot/mrrobust/master/")
mrdeps
```

* Fit IVW (with fixed effect standard errors)

```{stata}
mregger chdbeta ldlcbeta [aw=1/(chdse^2)] if sel1==1, ivw fe
```

* Fit MR-Egger reporting I<sup>2</sup><sub>GX</sub> statistic and heterogeneity Q-test

```{stata}
mregger chdbeta ldlcbeta [aw=1/(chdse^2)] if sel1==1, gxse(ldlcse) heterogi
```

* Simple plot of IVW and MR-Egger univariate estimates

```{stata}
mreggerplot chdbeta chdse ldlcbeta ldlcse if sel1==1
```

* Adding the modal and median estimates onto the plot

```{stata}
mreggerplot chdbeta chdse ldlcbeta ldlcse if sel1==1
mrmedian chdbeta chdse ldlcbeta ldlcse if sel1==1, weighted
addplot : function _b[beta]*x if sel1==1, ///
  range(0 0.5) lc(gs0) lp(shortdash) lw(vthin)
mrmodal chdbeta chdse ldlcbeta ldlcse if sel1==1, phi(.25)
addplot : function _b[beta]*x if sel1==1, ///
  range(0 0.5) lc(gs0) lp(longdash) ///
  legend(order(5 "Instruments"
               4 "95% CIs" 3 "MR-Egger" 2 "MR-Egger 95% CI" ///
               6 "Weighted median" 7 "Modal") ///
        rows(1) si(vsmall) symx(*.5))
```

* Fit and MVMR model with phenotypes LDL-c and HDL-c [@burgess-aje-2015].
```{stata}
mvmr chdbeta ldlcbeta hdlcbeta [aw=1/(chdse^2)] if sel1==1
```

* Additionally include a third phenotype -- triglycerides.
```{stata}
mvmr chdbeta ldlcbeta hdlcbeta tgbeta [aw=1/(chdse^2)] if sel1==1
```

* Report the Q<sub>A</sub> statistic for instrument validity and the conditional 
F-statistics for instrument strength for each phenotype [@sanderson-ije-2019; 
@sanderson-statsmed-2021].
```{stata}
mvmr chdbeta ldlcbeta hdlcbeta tgbeta [aw=1/(chdse^2)] if sel1==1, gxse(ldlcse hdlcse tgse)
```

* Fit an MVMR-Egger regression [@rees-statsmed-2017], orienting the model with 
respect to the first phenotype in the main *varlist*.
```{stata}
mrmvegger chdbeta ldlcbeta hdlcbeta tgbeta [aw=1/(chdse^2)] if sel1==1
```

We can also orient the model with respect to HDL-C.
```{stata}
mrmvegger chdbeta ldlcbeta hdlcbeta tgbeta [aw=1/(chdse^2)] if sel1==1, orient(2)
```

Or we can orient the model with respect to triglycerides.
```{stata}
mrmvegger chdbeta ldlcbeta hdlcbeta tgbeta [aw=1/(chdse^2)] if sel1==1,	orient(3)
```

## R session information for reproducibility

```{r}
library(sessioninfo)
session_info()
```

## Stata session information for reproducibility

```{stata}
about
ado describe mrrobust
```

## References
