---
title: "MR example for Supplementary Material"
# author: "Tom Palmer"
# date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
  word_document:
    reference_docx: "custom-reference.docx"
bibliography: references.bib 
---

```{r include=FALSE}
options(width = 100)

suppressPackageStartupMessages({
  if (!requireNamespace("tidyverse", quietly = TRUE)) install.packages("tidyverse")
  if (!requireNamespace("MendelianRandomization", quietly = TRUE)) install.packages("MendelianRandomization")
  if (!requireNamespace("MVMR", quietly = TRUE)) remotes::install_github("wspiller/mvmr")
  if (!requireNamespace("Statamarkdown", quietly = TRUE)) remotes::install_github("hemken/statamarkdown")
})
```

<!-- Move TOC to RHS -->
```{css toc-content, echo = FALSE}
#TOC {
  right: 270px;
  margin: 20px 0px 25px 0px;
}

.main-container {
    margin-left: 200px;
}
```

# Example R and Stata code to perform a multivariable Mendelian randomization (MVMR) analysis

## R code

* Read in the data

```{r}
suppressPackageStartupMessages({
  library(tidyverse)
  library(haven)
})
dat <- read_dta("https://raw.github.com/remlapmot/mrrobust/master/dodata.dta")
dat <- dat %>% filter(ldlcp2 < 1e-8)
```

### Example code using the MendelianRandomization package

* Install the package and load it into memory

```{r}
# install.packages("MendelianRandomization") # uncomment on first run
library(MendelianRandomization)
```

* Convert our data frame to the required class

```{r}
datfmt <- mr_mvinput(
  bx = as.matrix(cbind(dat$ldlcbeta, dat$hdlcbeta, dat$tgbeta)),
  bxse = as.matrix(cbind(dat$ldlcse, dat$hdlcse, dat$tgse)),
  by = dat$chdbeta,
  byse = dat$chdse,
  exposure = "exposure",
  outcome = "outcome",
  snps = "snp",
  effect_allele = dat$a1,
  other_allele = dat$a2,
  eaf = NA
)
```

* Fit an MVMR/MVIVW model

```{r}
mvivwfit <- mr_mvivw(datfmt)
mvivwfit
```

* Fit an MVMR-Egger model

```{r}
mvmreggerfit <- mr_mvegger(datfmt)
mvmreggerfit
```

### Example code using the MVMR package

* Install the package and load it into memory

```{r}
# remotes::install_github("wspiller/mvmr") # uncomment on first run
library(MVMR)
```

* Create a data object of the required structure

```{r}
r_input <- format_mvmr(
  BXGs = dat[,c("ldlcbeta","hdlcbeta","tgbeta")],
  BYG = dat$chdbeta,
  seBXGs = dat[,c("ldlcse","hdlcse","tgse")],
  seBYG = dat$chdse,
  RSID = dat$rsid
)
```

* Fit an MVMR model

```{r}
mvmrfit <- ivw_mvmr(r_input)
```

* Heterogeneity statistic

```{r}
strength_mvmr(r_input)
```

## Stata code

* Load the Statamarkdown package to enable Stata code chunks in an R Markdown file

```{r}
# remotes::install_github("hemken/statamarkdown") # uncomment on first run
library(Statamarkdown)
```

* Read in the data and create an indicator variable to select observations with *p*-value between the genotype and LDL-C < 10<sup>-8</sup>

```{stata, collectcode=TRUE}
use https://raw.github.com/remlapmot/mrrobust/master/dodata, clear
gen byte sel1 = (ldlcp2 < 1e-8)
```

### Example code using the mrrobust package

* Install the mrrobust package using the github package

```{stata results=FALSE}
// output suppressed
net install github, from("https://haghish.github.io/github/") replace
gitget mrrobust
```

* Fit and MVMR model with phenotypes LDL-c and HDL-c [@burgess-aje-2015].
```{stata}
mvmr chdbeta ldlcbeta hdlcbeta [aw=1/(chdse^2)] if sel1==1
```

* Additionally include a third phenotype -- triglycerides.
```{stata}
mvmr chdbeta ldlcbeta hdlcbeta tgbeta [aw=1/(chdse^2)] if sel1==1
```

* Report the Q<sub>A</sub> statistic for instrument validity and the conditional 
F-statistics for instrument strength for each phenotype [@sanderson-ije-2019; 
@sanderson-biorxiv-2020].
```{stata}
mvmr chdbeta ldlcbeta hdlcbeta tgbeta [aw=1/(chdse^2)] if sel1==1, ///
	gxse(ldlcse hdlcse tgse)
```

* Fit an MVMR-Egger regression [@rees-statsmed-2017], orienting the model with 
respect to the first phenotype in the main *varlist*.
```{stata}
mrmvegger chdbeta ldlcbeta hdlcbeta tgbeta [aw=1/(chdse^2)] if sel1==1
```

We can also orient the model with respect to HDL-C.
```{stata}
mrmvegger chdbeta ldlcbeta hdlcbeta tgbeta [aw=1/(chdse^2)] if sel1==1, ///
	orient(2)
```

Or we can orient the model with respect to triglycerides.
```{stata}
mrmvegger chdbeta ldlcbeta hdlcbeta tgbeta [aw=1/(chdse^2)] if sel1==1, ///
	orient(3)
```

## R session information for reproducibility

```{r}
if (!requireNamespace("sessioninfo", quietly = TRUE)) install.packages("sessioninfo")
sessioninfo::session_info()
```

## Stata session information for reproducibility

```{stata}
di c(version)
github list mrrobust
```

## References
